Parametricity for Primitive Nested Types and GADTs

### 2020-12-20 free-theorems-nesteds/20-mscs.tex
added GADT section to be revised

  \overline \alpha.H)\overline \beta \;F)}\] in $\rho$ is trivial.

@@

>>>>

  \overline \alpha.H)\overline \beta \;F)}\] in $\rho$ is trivial.\qed

@@

Generalized algebraic datatypes (GADTs) generalize nested types by...

For example, ...

A natural next step in this line of work is to extend our
parametricity results to GADTs.  It was shown in~\cite{jg08} that (the
data of) GADTs can be represented using object-level left Kan
extensions (over discrete categories), and further shown
in~\cite{jp19} that adding a carefully designed left Kan extension
construct to a calculus supporting primitive nested types preserves
the cocontinuity needed for primitive GADTs to have well-defined,
properly functorial interpretations.

This suggests adding GADTs to our calculus by extending
Definition~\ref{def:wftypes} with such object-level constructs, and
extending Figure~\ref{fig:terms} with categorically inspired
constructs to introduce and eliminate terms of these types, along the
same entirely standard lines as we did there for products, coproducts,
and fixpoints.  In this section we outline how our model from
Section~\ref{sec:term-interp} can be extended to a parametric model
when some classes of GADTs are incorporated as primitives in this
way. However, as we argue in~???? below, we suspect this is not
possible in general.

HERE!!

To show when this can be done, as well as to give a sense of what goes
wrong in the general case, we begin by recalling that GADTs are
naturally interpreted as initial algebras of left Kan extensions, as
suggested in discrete~\cite{jg08} and

Functorial semantics in~\cite{jp19}.

introducing a new
type, $\Lan$, to represent left Kan extensions.  Indeed,

To begin with, we add the following type formation rule for $\Lan$
types to our type system.
\[
\AXC{$\Gamma; \Phi,\ol\alpha^0 \vdash F$}
\AXC{$\ol{\Gamma;\ol\alpha^0 \vdash K}$}
\AXC{$\ol{\Gamma;\Phi \vdash A}$}
\TIC{$\Gamma;\Phi \vdash (\Lan^{\ol{\alpha^0}}_{\ol K} F) \ol A$}
\DisplayProof
\]
Notice that the length of $\ol K$ coincides with that of $\ol A$.
Introducing the new type also requires the following additional clause in the inductive definition of higher-order substitution.
\[
((\Lan^{\ol{\alpha}}_{\ol K}\,F) \ol A)[\phi :=_{\ol{\beta}} G]
  = (\Lan^{\ol\alpha}_{\ol K}\,F[\phi :=_{\ol{\beta}} G])\, \ol{A[\phi :=_{\ol{\beta}} G]}
\]

Before we extend the term calculus with the term introduction and elimination rules for $\Lan$ types,
it is necessary to generalize $\Nat$ types to bind type constructor variables of arbitrary arity.
So, the formation rule for $\Nat$ types will be modified as
\[
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi  \vdash G$}
\BIC{$\Gamma;\emptyset \vdash \Nat^\Phi F \,G$}
\DisplayProof
\]
and the term introduction and elimination rules will be
\[
\AXC{$\Gamma; \Phi \vdash F$}
\AXC{$\Gamma; \Phi \vdash G$}
\AXC{$\Gamma; \Phi \,|\, \Delta, x : F \vdash t: G$} 
\TIC{$\Gamma; \emptyset
  \,|\, \Delta \vdash L_\Phi x.t : \Nat^\Phi \,F \,G$}
\DisplayProof
\]
and
\[
\AXC{$\ol{\Gamma;\Phi,\ol\beta \vdash K}$}
\AXC{$\Gamma; \emptyset
  \,|\, \Delta \vdash t : \Nat^{\ol\psi} \,F \,G$}
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\psi :=_{\ol\beta} K}]$}
\TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
  G[\overline{\psi :=_{\ol\beta} K}]$}
\DisplayProof
\]
respectively.
It is necessary, of course, to suitably extend the semantic interpretations of $\Nat$ types and terms.
The type constructor variables bound by $\Nat$ will be interpreted as either functors
or relation transformers.
The set and relational interpretations for $\Nat$ types are defined as follows.
\begin{align*}
\setsem{\Gamma; \emptyset \vdash \Nat^\Phi\,F\,G}\rho
&= \{\eta : \lambda \ol{K}. \,\setsem{\Gamma; \Phi \vdash F}\rho[\Phi := \ol K] 
  \Rightarrow \lambda \ol{K}.\,\setsem{\Gamma; \Phi \vdash G}\rho[\Phi := \ol K]~|~\\ 
&\hspace{0.3in} \forall \ol{K = (K^1, K^2, K^*) : RT_k}.\\
&\hspace{0.4in} (\eta_{\overline{K^1}}, \eta_{\overline{K^2}})
: \setsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\Phi := \ol K]
\rightarrow \setsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\Phi := \ol K] \} \\
%%%
\relsem{\Gamma; \emptyset \vdash \Nat^\Phi \,F\,G}\rho
&= \{\eta : \lambda \ol{K}.\,\relsem{\Gamma; \Phi \vdash F}\rho[\ol{\Phi := K}]
\Rightarrow \lambda \ol{K}. \,\relsem{\Gamma; \Phi \vdash G}\rho[\ol{\phi := K}]\}\\
&=
  \{(\eta_1,\eta_2) \in \setsem{\Gamma; \emptyset
    \vdash \Nat^\Phi
    \,F\,G} (\pi_1 \rho) \times \setsem{ 
    \Gamma;\emptyset
    \vdash \Nat^\Phi \,F\,G} (\pi_2
  \rho)~|~\\
& \hspace{0.3in} \forall \ol{K = (K^1, K^2, K^*) : RT_{k}}.\\
& \hspace{0.4in} ((\eta_1)_{\ol{K^1}},(\eta_2)_{\ol{K^2}}) \in
  (\relsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi := K}])^{\relsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K}]} \}\\  
\end{align*}

We can now give the term introduction and elimination rules for $\Lan$, which are
\[
\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
\AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
\BIC{$\Gamma;\emptyset~|~\Delta \vdash \int_{\ol K,F} : \Nat^{\Phi}\, F\, (\Lan^{\ol\alpha}_{\ol{K}}\,  F) \ol{K}$}
\DisplayProof
\]
and
\[
\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\ol\alpha}\,
  F\;G[\ol{\beta := K}]$}
\UIC{$\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F \eta :
\Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F)\ol\beta\, G$}
\DisplayProof
\]
respectively.
Notice that, in these rules, we make essential use of the extended $\Nat$ types.

To give the semantics for $\Lan$ types we follow~\cite{jp19}
and use the fundamental idea that these have to be interpreted as left Kan extensions.
We give the definition of left Kan extension and some useful notation in the following definition.

\begin{defi}
Let $F : \set^k \to \set$ and $K : \set^k \to \set^h$.
Then, the left Kan extension of $F$ along $\ol K$ is given by a functor $\Lan_K\,F : \set^h \to \set$
and a natural transformation $\epsilon : F \to (\Lan_K\,F) \circ K$
which are initial among such pairs of functor and natural transformations, i.e.,
such that for every $L : \set^h \to \set$ and $\eta :  F \to L \circ K$
there exists a unique natural transformation $\mu : \Lan_K\,F \to L$
such that $(\mu K) \circ \epsilon = \eta$, as shown in the following diagram.
\[
\begin{tikzcd}[row sep = huge]
\set^k
\ar[rr, "{F}"{name=Fa}, ""{name=F, below}]
\ar[rd, "{K}"']
&& \set \\
& \set^h
\ar[Rightarrow, bend right = 10, from=F, "{\epsilon}" description]
\ar[ur, bend left, "{\Lan_K\,F}"{name=Lan, description}]
\ar[ur, bend right, "{L}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 90, from=Fa, to=Lr, "{\eta}"]
\ar[Rightarrow, from=Lan, to=L, "{\mu}"]
\end{tikzcd}
\]
Of course, we can similarly define the left Kan extension for functors on $\rel$.
\end{defi}

One might try to restrict the semantic interpretation of $\Lan$ types to a subset of the set produced by the left Kan extension.
We show in the following proposition that this is not possible, if we want such interpretation to include the interpretation of the terms given by the term-introduction rule.

\begin{prop}\label{prop:full-lan}
Let $F : \set^h \to \set$ and $\ol{K : \set^h \to \set}$ be functors.
Assume that $L : \set^k \to \set$ is a functor such that the following properties hold:
\begin{enumerate}
\item for all $\ol{A : \set}$, we have that $L \ol{A} \subseteq (\textit{Lan}_{\ol K}\, F)\ol{A}$
\item for all $\ol{f : A \to B}$ and $\ol{x \in A}$, we have that $(\textit{Lan}_{\ol K}\, F)\ol{f}|_{L \ol{A}}\, x \in L \ol{B}$
\item for all $\ol{A : \set}$ and $y \in F \ol{A}$, we have that $\epsilon_{\ol{A}} y \in L ( \ol{K \ol{A}} )$
\end{enumerate}
Then $L = \textit{Lan}_{\ol K}\, F$.
\end{prop}
\begin{proof}
For every $\ol{A : \set}$ and $z \in (\textit{Lan}_{\ol K}\, F)\ol{A}$
there exist $\ol{S : \set_0}$, $\ol{f : K \ol{S} \to A}$ and $w \in F \ol{S}$ such that $z = \iota_{\ol{S}, \ol{f}} w$.
Thus, $z = \iota_{\ol{S}, \ol{f}} w = (\textit{Lan}_{\ol K}\, F) \ol{f} (\epsilon_{\ol{S}} w)$;
by the third assumption we have that $\epsilon_{\ol{S}} w \in L ( \ol{K \ol{S}} )$
and by the second assumption we have that $z \in L \ol{A}$.
We can than conclude that $L \ol{A} = (\textit{Lan}_{\ol K}\, F)\ol{A}$.
\end{proof}

Thus, we are forced to interpret $\Lan$ types as whole left Kan extensions, without any cutting down.
The set interpretation of the $\Lan$ type then is
\[
  \setsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}\rho
  = (\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}
  \rho[\ol{\alpha := \_}]}}\,
  \setsem{\Gamma; \Phi, \ol\alpha \vdash F}
  \rho[\ol{\alpha := \_}])\,
  \ol{\setsem{\Gamma; \Phi \vdash A}\rho}
\]
and the relational interpretation is
\[
  \relsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}\rho
  = (\textit{Lan}_{\ol{\relsem{\Gamma; \ol\alpha \vdash
          K}\rho[\ol{\alpha := \_}]}}\, \relsem{\Gamma; \Phi, \ol\alpha
      \vdash F}\rho[\ol{\alpha := \_}])\, \ol{\relsem{\Gamma; \Phi
        \vdash A}\rho}
\]
We also need to define the functorial action of the interpretation on morphisms.
Let $f : \rho \to \rho'$ be a morphism of set environments.
Then
\[
\setsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}f
: \setsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}\rho
\to \setsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}\,F)\ol{A}}\rho'
\]
is defined as
\begin{multline*}
\big(\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]\big)\,
    \ol{\setsem{\Gamma; \Phi \vdash A}\rho'} \\
\circ\;
  \big(\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}]\big)\,
    \ol{\setsem{\Gamma; \Phi \vdash A}f}
\end{multline*}
or, equivalently by naturality,
\begin{multline*}
\big(\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := \_}]\big)\,
    \ol{\setsem{\Gamma; \Phi \vdash A}f} \\
\circ\;
  \big(\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]\big)\,
    \ol{\setsem{\Gamma; \Phi \vdash A}\rho}
\end{multline*}
The functorial action for the relational interpretation is defined analogously.

The left Kan extension can be expressed via a colimit formula~\cite{rie16} as
\[
	(\Lan_K\,F) \ol{A} = \colim{\ol{S} : \set^k_0, \ol{f : K\ol{S} \to A}}{F \ol{S}}
\]
for any $\ol{A} : \set^h$, and where $\set_0$ is the category finite sets
(i.e., the full subcategory of $\omega$-presentable objects in $\set$).
For any $\ol{S} : \set^k_0$ and $\ol{f : K\ol{S} \to A}$, let
$j_{\ol{S}, \ol{f}} : F \ol{S} \to (\Lan_K\,F) \ol{A}$
be the morphism indexed by $\ol{S}$ and $\ol{f}$ of the cocone into the colimit.
Moreover, for any $h$-tuple of functions $\ol{g : A \to B}$,
the functorial action $(\Lan_K\,F) \ol{g}$ is the unique function $(\Lan_K\,F) \ol{A} \to (\Lan_K\,F) \ol{B}$
such that, for every $\ol{S} : \set^k_0$ and $\ol{f : K\ol{S} \to A}$,
we have that $((\Lan_K\,F) \ol{g}) \circ j_{\ol{S}, \ol{f}} = j_{\ol{S}, \ol{g \circ f}}$.
We can similarly express via a colimit formula the left Kan extension of functors on $\rel$,
and in that case we call the cocone over it $\iota$.

We define the semantics for the terms from the introduction and elimination rules for $\Lan$ types as, respectively,
\begin{multline*}
\setsem{\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t :
  (\Lan^{\ol\alpha}_{\ol{K}}\, F)\, \ol{K[\ol{\alpha := A}]}}\rho\;d \\
= j_{\ol{\setsem{\Gamma;\Phi \vdash A}\rho},
  \ol{\id_{\setsem{\Gamma; \Phi \vdash K[\ol{\alpha := A}]}\rho}}}\,
(\setsem{\Gamma; \Phi~|~\Delta \vdash t : F[\ol{\alpha := A}]} \rho\, d)
\end{multline*}
and
\[
\setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F \eta 
: \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F) \ol \beta\; G}\rho
= \lambda \ol{N} \, \ol{B}.\, (\mu_{\ol{N}})_{\ol{B}}
\]
where
  \[\mu_{\ol{N}} : \Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho
  {[\ol{\alpha := \_} ]}}} \setsem{\Gamma; \Phi, \ol\alpha \vdash
    F}\rho {[\Phi, \ol \alpha := \ol N, \_ ]} \to \setsem{\Gamma;
    \Phi, \ol\beta \vdash G} \rho[\Phi, \ol \beta := \ol N, \_]\]
  is the unique natural transformation such that
\begin{equation}\label{eq:defn-of-mu-sub-N}
    (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta : \Nat^{\Phi,
    \ol\alpha} \, F \,\, G[\ol{\beta := K}]}\rho \, d)_{\ol{N}} =
(\mu_{\ol{N}} \, \ol{\setsem{\Gamma; \ol\alpha \vdash K}
  \rho[\ol{\alpha := \_}]}) \circ \epsilon_{\ol{N}}
  \end{equation}
as given by the universal property of the left Kan extension.
Here, $\epsilon_{\ol{N}}$ is the natural transformation associated
with the left Kan extension, as shown in the following diagram.
\[\begin{tikzcd} [column sep = large, row sep = large]
\set^{|\ol \alpha|}
\ar[rr, "\setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho {[\Phi, \ol
      \alpha := \ol N, \_ ]}"{name=F}] 
\ar[dr, "\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := \_} ]}}"{left}] && \set \\ 
&\set^{|\ol K|} \ar[ur, "\Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := \_} ]}}}\,
\setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha :=
    \ol N, \_ ]}"{right}]
\ar[d, Rightarrow, shorten <= 2mm, shorten >= 1mm, 
          from=F, to=2-2,  "\epsilon_{\ol{N}}"{}]
\end{tikzcd}\]
The relational interpretations are define analogously.

Having defined the syntax and both the set and relational semantics for $\Lan$ types and terms,
we need to check that these definitions yield a parametric model.
The first step would be to extend Lemma~\ref{lem:rel-transf-morph} to $\Lan$ types,
by proving a new clause of the inductive step of the proof.
Unfortunately, it turns out that, in general, this is not possible, as the following example shows.

\begin{exa}
We provide a counterexample to Lemma~\ref{lem:rel-transf-morph} for $\Lan$ types.
Consider the type $\emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet}\, \onet) \alpha$.
Then, using the colimit formula for left Kan extensions, we have
\[
\begin{array}{rl}
\relsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet}\, \onet) \alpha } \rho [\alpha := (1, 0, 0)]
&= (\textit{Lan}_{\Eq_1}\, \Eq_1)(1, 0, 0) \\
&= \colim{f : \Eq_1 \to (1, 0, 0)}{\Eq_1} \\
&= (0, 0, 0)
\end{array}
\]
because there exists no morphism $f : \Eq_1 \to (1, 0, 0)$ in $\rel$.
But, considering the set interpretation on the first projection of the relation environment, we have
\[
\begin{array}{rl}
\setsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet}\, \onet) \alpha } ( \pi_1 ( \rho [\alpha := (1, 0, 0)] ) )
&= \setsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet}\, \onet) \alpha } (\pi_1 \rho) [\alpha := 1] \\
&= (\textit{Lan}_{1}\, 1) 1 \\
&= \colim{1 \to 1}{1} \\
&= 1
\end{array}
\]
because there exist a unique arrow $1 \to 1$ in $\set$, which is the identity on $1$.
\end{exa}

The issue lies in the fact that the functor along which we extend contains constants (such as $\onet$).
Indeed, if we assume that $\ol{\emptyset; \overline{\alpha} \vdash K}$ are polynomial functors not containing constants,
then we can prove that the relational interpretation of
$\Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A}$
is fibered over its set interpretation, as proven in the following proposition.

\begin{prop}
Assume that the types in $\ol{K}$ are polynomial and contain no constants
(i.e., they are formed only by the functorial type variables $\ol\alpha$, $+$ and $\times$).
Then
\[
\pi_i (
\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
)
=
\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i \rho)
\]
for every relation environment $\rho$ and $i = 1, 2$.
\end{prop}
\begin{proof}
Notice that, by the colimit formula,
\[
\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
=
\colim{\ol{R}, \ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}}{
\relsem{ \Gamma; \Phi, \ol\alpha \vdash F } \rho[\ol{\alpha := R}]
}
\]
(where we call $\iota$ the cocone of the colimit)
and thus
\[
\pi_i (
\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
)
=
\colim{\ol{R : \rel_0}, \ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}}{
\relsem{ \Gamma; \Phi, \ol\alpha \vdash F } (\pi_i \rho) [\ol{\alpha := \pi_i R}]}
\]
(where the cocone of the colimit is $\pi_i \iota$),
whereas, again by the colimit formula,
\[
\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i \rho)
=
\colim{\ol{S : \set_0}, \ol{f : \setsem{ \emptyset; \ol\alpha \vdash K } (\pi_i \rho)[\ol{\alpha := S}] \to \setsem{ \Gamma; \Phi \vdash A } (\pi_i \rho)}}{
\setsem{ \Gamma; \Phi, \ol\alpha \vdash F } (\pi_i \rho) [\ol{\alpha := S}]}
\]
(where we call $j$ the cocone of the colimit).

We now prove that
\[
\pi_1 (
\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
)
=
\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_1 \rho)
\]
by providing a pair of inverse functions
(the proof for the second projection is analogous).
Define
\[
h :
\pi_1 (
\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
)
\to
\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_1 \rho)
\]
as the unique function given by the universal property of the colimit such that, for any $\ol{R : \rel_0}$ and $\ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}$, we have that $h \circ (\pi_1 \iota_{\ol{R}, \ol{m}}) = j_{\ol{\pi_1 R}, \ol{\pi_1 m}}$.

Observe that, since $K$ does not contain constants, then $\setsem{ \emptyset; \ol\alpha \vdash K } (\pi_2 \rho) [\ol{\alpha := 0}] = 0$.
For any
\[
\ol{S : \set_0}
\qquad \text{and} \qquad
\ol{f : \setsem{ \emptyset; \ol\alpha \vdash K } (\pi_1 \rho)[\ol{\alpha := S}] \to \setsem{ \Gamma; \Phi \vdash A } (\pi_1 \rho)}
\]
consider the relations $\ol{(S, 0, 0)}$ and the morphisms $\ol{(f, !) : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := (S, 0, 0)}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}$ (where $!$ is the unique function from the empty set).
Then, define
\[
k :
\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_1 \rho)
\to
\pi_1 (
\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
)
\]
as the unique function given by the universal property of the colimit such that, for any $\ol{S : \set_0}$ and $\ol{f : \setsem{ \emptyset; \ol\alpha \vdash K } (\pi_1 \rho)[\ol{\alpha := S}] \to \setsem{ \Gamma; \Phi \vdash A } (\pi_1 \rho)}$, we have that $k \circ j_{\ol{S}, \ol{f}} = \pi_1 \iota_{\ol{(S, 0, 0)}, \ol{(f, !)}}$.

We now have to show that $h$ and $k$ are mutually inverse.
First, we prove that $h \circ k = \id$.
Indeed, observe that
\[
h \circ k \circ j_{\ol{S}, \ol{f}}
= h \circ (\pi_1 \iota_{\ol{(S, 0, 0)}, \ol{(f, !)}})
= j_{\ol{\pi_1 (S, 0, 0)}, \ol{\pi_1 (f, !)}}
= j_{\ol{S}, \ol{f}}
\]
for every $\ol{S : \set_0}$ and
$\ol{f : \setsem{ \emptyset; \ol\alpha \vdash K } (\pi_1 \rho)[\ol{\alpha := S}] \to \setsem{ \Gamma; \Phi \vdash A } (\pi_1 \rho)}$
and thus $h \circ k = \id$.
Secondly, we prove that $k \circ h = \id$.
Indeed, observe that
\[
k \circ h \circ (\pi_1 \iota_{\ol{R}, \ol{m}})
= k \circ j_{\ol{\pi_1 R}, \ol{\pi_1 m}}
= \pi_1 \iota_{\ol{(\pi_1 R, 0, 0)}, \ol{(\pi_1 m, !)}}
\]
for every $\ol{R : \rel_0}$ and $\ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}$.
Moreover, consider the morphisms $\ol{(\id_{\pi_1 R}, !) : (\pi_1 R, 0, 0) \to R}$ and observe that $m \circ \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := (\id_{\pi_1 R}, !)}] = (\pi_1 m, !)$, implying that $\iota_{\ol{R}, \ol{m}} = \iota_{\ol{(\pi_1 R, 0, 0)}, \ol{(\pi_1 m, !)}}$.
Then $k \circ h \circ (\pi_1 \iota_{\ol{R}, \ol{m}}) = \pi_1 \iota_{\ol{R}, \ol{m}}$ for every $R$ and $m$, and thus $h \circ k = \id$.

We can also prove that
\[
\pi_i (
\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f
)
=
\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i f)
\]
for every morphism of relation environments $f : \rho \to \rho'$ and $i = 1, 2$.
Let $\iota$ be the cocone for the colimit $\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho$,
$j$ the cocone for the colimit $\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i\rho)$
and $h$ and $k$ the isomorphisms showing that
$\pi_i(\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho) = \setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i\rho)$;
analogously, let $\iota'$, $j'$, $h'$ and $k'$ be as above but for interpretations in $\rho'$ in place of $\rho$.
Then $\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f$
is the unique morphism of relations such that, for any $\ol{R : \rel_0}$ and
$\ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}$,
we have that
\[
(\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f) \circ \iota_{\ol{R}, \ol{m}}
= \iota'_{\ol{R}, \ol{\relsem{ \Gamma; \Phi \vdash A } f \circ m}}
\]
but then, by projecting the $i$-th component, we have that
\[
\pi_i (\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f) \circ \pi_i \iota_{\ol{R}, \ol{m}}
= \pi_i \iota'_{\ol{R}, \ol{\relsem{ \Gamma; \Phi \vdash A } f \circ m}}
\]
and thus, using the facts that $h \circ (\pi_i \iota_{\ol{R}, \ol{m}}) = j_{\ol{\pi_i R}, \ol{\pi_i m}}$
and $h' \circ (\pi_i \iota'_{\ol{R}, \ol{m}}) = j'_{\ol{\pi_i R}, \ol{\pi_i m}}$ from above,
we get
\[
\pi_i (\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f) \circ k \circ j_{\ol{\pi_i R}, \ol{\pi_i m}}
= k' \circ j'_{\ol{\pi_i R}, \ol{\relsem{ \Gamma; \Phi \vdash A } (\pi_i f) \circ (\pi_i m)}}
\]
i.e., $\pi_i (\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } f)$
satisfy the universal property of $\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i f)$,
and thus they are equal (up to natural isomorphism).
\end{proof}

The requirement that the $\ol{K}$s be polynomial functors without constants is quite restrictive,
as it precludes the possibility of expressing GADTs used in practice, such as the following type $\mathtt{Expr}$ of typed expressions.
\[
\begin{array}{l}
\mathtt{data\; Expr \;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{const\;:\; Int \to Expr\; Int}\\
\hspace*{0.4in}\mathtt{is\_zero\;:\; Expr\; Int \rightarrow Expr\; Bool}\\
\hspace*{0.4in}\mathtt{if\;:\; Expr\; Bool \to Expr\; A \to Expr\; A \to Expr\; A}
\end{array}
\]

{\color{red} This is not quite right.}
Even when extending along polynomials containing no constants, it is unclear how to prove the Identity Extension Lemma for $\Lan$ types, or if it holds at all.
In order to satisfy the identity extension lemma in the case of $\Nat$ types, for example, it is necessary to cut down the semantic interpretation by requiring that the natural transformations in the set interpretation preserve equalities.
The impossibility of restricting the semantic interpretation of $\Lan$, as proved by Proposition~\ref{prop:full-lan}, suggests that the Identity Extension Lemma might indeed not hold for a type system that features such types.

We must then conclude that it is not possible to use functorial semantics to provide a parametric model for $\Lan$ types.
